The lattice of acyclic pipe dreams

### 2022-11-29 acyclicFacets/draft/latticePipeDreams.tex
minor changes

%Two pipe dreams are related by a \defn{flip} if they only differ by the position of a crossing and a contact (in particular, they define the same permutation).
A contact~$c$ is \defn{flippable} if the two pipes passing through contact~$c$ have a crossing~$x$.
The \defn{flip} exchanges the contact~$c$ with the crossing~$x$.
%See \cref{fig:pipeDreams}.
The flip is \defn{increasing} if the contact~$c$ is weakly southwest of the crossing~$x$.
For example, \cref{fig:pipeDreams} illustrates an increasing from the left pipe dream to the right pipe dream.
%
The \defn{increasing flip graph} is the graph of increasing flips on~$\pipeDreams(\omega)$.
It is clearly a directed acyclic graph, and it has a unique source and a unique sink \cite{PilaudPocchiola}, called the \defn{greedy} and \defn{antigreedy} pipe dreams, and denoted~$\greedyPipeDream$ and~$\antiGreedyPipeDream$.
\vincent{In \cite{PilaudStump-ELlabeling}, these are called positive and negative greedy facets. Also the notations are different...}
The \defn{increasing flip poset} is the reflexive and transitive closure of the increasing flip graph on~$\pipeDreams(\omega)$.

@@

\begin{remark}
\label{rem:pipeDreamQuotient}
The increasing flip poset on~$\pipeDreams(\omega)$ is the transitive closure of the increasing flip graph on~$\pipeDreams(\omega)$.
Observe that the two natural ways to restrict to the acyclic pipe dreams of~$\acyclicPipeDreams(\omega)$ (restrict either the flip graph or the flip poset) do not coincide in general.
Namely, the transitive closure of the subgraph of the increasing flip graph induced by~$\acyclicPipeDreams(\omega)$ may have strictly less relations than the subposet of the increasing flip poset induced by~$\acyclicPipeDreams(\omega)$.
\cref{fig:counterExampleRestrictionIncreasingFlipPoset} illustrates two acyclic pipe dreams of~$\acyclicPipeDreams(126543)$ connected by a sequence of increasing flips in~$\pipeDreams(126543)$ but by no sequence of increasing flips in~$\acyclicPipeDreams(126543)$.
%	* type A5,
%	* Q = [1, 2, 3, 4, 5, 1, 2, 3, 4, 1, 2, 3, 1, 2, 1] 
%	* w = [1, 2, 3, 1, 2, 1]
%	* flips: 
%	(0, 2, 3, 4, 5, 6, 7, 8, 12)
%	(2, 3, 4, 5, 6, 7, 8, 10, 12) 
%	(3, 4, 5, 6, 7, 8, 10, 11, 12)
%	(3, 4, 5, 7, 8, 10, 11, 12, 14)

@@

\begin{proposition}
\label{prop:sweepingAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed by sweeping the triangular shape in any perturbation of the northwest direction and inserting a crossing when sweeping a vertex~$v$ of the grid where pipe~$i$ arrives horizontally and pipe~$j$ arrives vertically if and only if
\begin{itemize}
\item $i < j$ and~$\omega^{-1}(i) > \omega^{-1}(j)$, and 
\item $\pi^{-1}(i) > \pi^{-1}(j)$ or vertex~$v$ lies in column~$\omega^{-1}(j)$.
\end{itemize}
See \cref{fig:sweepingAlgorithm}.
\vincent{todo}

\begin{figure}
%	\centerline{\includegraphics[scale=.9]{sweeping}}
	\caption{Sweeping algorithm for the permutations~$\pi = 513264$ and~$\omega = 561324$.}
	\label{fig:sweepingAlgorithm}
\end{figure}

@@

\begin{proposition}
\label{prop:insertionAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed starting from the empty triangular shape and inserting the pipes~$\pi(1), \dots, \pi(n)$ one by one in the order of the permutation~$\pi$ as northwest as possible.
More precisely, at step~$t$, we consider the set~$X_t$ of free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$, and we insert a pipe starting at row~$\pi(t)$, ending at column~$\omega^{-1}(\pi(t))$, covering each elbow of~$X_t$, and adding a new free southeast elbow in between any two consecutive elbows of~$X_t$
See \cref{fig:insertionAlgorithm}.

@@

\begin{lemma}
\label{lem:rectangleInsertionAlgorithm}
At each step~$t$ of the insertion algorithm, we have
\[
\#X_t = \# \bigset{s < t}{\pi(s) < \pi(t) \text{ and } \omega^{-1}\big( \pi(s) \big) < \omega^{-1}\big( \pi(t) \big)}.
\]
\vincent{Note that this is independent of~$\pi$... This can be seen from \cref{lem:interval}.}
\end{lemma}

>>>>

%Two pipe dreams are related by a \defn{flip} if they only differ by the position of a crossing and a contact (in particular, they define the same permutation).
A contact~$c$ is \defn{flippable} if the two pipes passing through contact~$c$ have a crossing~$x$.
The \defn{flip} exchanges the contact~$c$ with the crossing~$x$.
%See \cref{fig:pipeDreams}.
The flip is \defn{increasing} if the contact~$c$ is weakly southwest of the crossing~$x$.
For example, \cref{fig:pipeDreams} illustrates an increasing flip from the left pipe dream to the right pipe dream.
%
The \defn{increasing flip graph} is the graph of increasing flips on~$\pipeDreams(\omega)$.
It is clearly a directed acyclic graph, and it has a unique source and a unique sink \cite{PilaudPocchiola}, called the \defn{greedy} and \defn{antigreedy} pipe dreams, and denoted~$\greedyPipeDream$ and~$\antiGreedyPipeDream$.
\vincent{In \cite{PilaudStump-ELlabeling}, these are called positive and negative greedy facets. Also the notations are different...}
The \defn{increasing flip poset} is the reflexive and transitive closure of the increasing flip graph on~$\pipeDreams(\omega)$.

@@

\begin{remark}
\label{rem:pipeDreamQuotient}
The increasing flip poset on~$\pipeDreams(\omega)$ is the transitive closure of the increasing flip graph on~$\pipeDreams(\omega)$.
Observe that the two natural ways to restrict to the acyclic pipe dreams of~$\acyclicPipeDreams(\omega)$ (restrict either the flip graph or the flip poset) do not coincide in general.
Namely, the transitive closure of the subgraph of the increasing flip graph induced by~$\acyclicPipeDreams(\omega)$ may have strictly less relations than the subposet of the increasing flip poset induced by~$\acyclicPipeDreams(\omega)$.
\cref{fig:counterExampleRestrictionIncreasingFlipPoset} illustrates two acyclic pipe dreams of~$\acyclicPipeDreams(126543)$ connected by a sequence of increasing flips in~$\pipeDreams(126543)$ but by no sequence of increasing flips in~$\acyclicPipeDreams(126543)$.
This example is minimal.
%	* type A5,
%	* Q = [1, 2, 3, 4, 5, 1, 2, 3, 4, 1, 2, 3, 1, 2, 1] 
%	* w = [1, 2, 3, 1, 2, 1]
%	* flips: 
%	(0, 2, 3, 4, 5, 6, 7, 8, 12)
%	(2, 3, 4, 5, 6, 7, 8, 10, 12) 
%	(3, 4, 5, 6, 7, 8, 10, 11, 12)
%	(3, 4, 5, 7, 8, 10, 11, 12, 14)

@@

\begin{proposition}
\label{prop:sweepingAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed by sweeping the triangular shape in any perturbation of the northwest direction and inserting a crossing when sweeping a vertex~$v$ of the grid where pipe~$i$ arrives horizontally and pipe~$j$ arrives vertically if and only if
\begin{itemize}
\item $i < j$ and~$\omega^{-1}(i) > \omega^{-1}(j)$, and 
\item $\pi^{-1}(i) > \pi^{-1}(j)$ or vertex~$v$ lies in column~$\omega^{-1}(j)$.
\end{itemize}
See \cref{fig:sweepingAlgorithm}.
%
\begin{figure}
	\centerline{\includegraphics[scale=.9]{sweeping}}
	\caption{Sweeping algorithm for the permutations~$\pi = 513264$ and~$\omega = 561324$.}
	\label{fig:sweepingAlgorithm}
\end{figure}

@@

\begin{proposition}
\label{prop:insertionAlgorithm}
For any permutations~$\pi,\omega \in \fS_n$ such that~$\pi < \omega$ in weak order, the unique pipe dream~$P \in \acyclicPipeDreams(\omega)$ such that~$\pi \in \linearExtensions(P)$ can be constructed starting from the empty triangular shape and inserting the pipes~$\pi(1), \dots, \pi(n)$ one by one in the order of the permutation~$\pi$ as northwest as possible.
More precisely, at step~$t$, we insert a pipe starting at row~$\pi(t)$, ending at column~$\omega^{-1}(\pi(t))$, and whose northeast elbows are precisely covering the free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$.
%More precisely, at step~$t$, we consider the set~$X_t$ of free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$, and we insert a pipe starting at row~$\pi(t)$, ending at column~$\omega^{-1}(\pi(t))$, and whose northeast elbows are precisely covering the elbows of~$X_t$, and adding a new free southeast elbow in between any two consecutive elbows of~$X_t$
See \cref{fig:insertionAlgorithm}.

@@

\begin{lemma}
\label{lem:rectangleInsertionAlgorithm}
At each step~$t$ of the insertion algorithm, the number of free southeast elbows in the rectangle~$[\pi(t)] \times [\omega^{-1}(\pi(t))]$ is
\[
\# \bigset{s < t}{\pi(s) < \pi(t) \text{ and } \omega^{-1}\big( \pi(s) \big) < \omega^{-1}\big( \pi(t) \big)}.
\]
\vincent{Note that this is independent of~$\pi$... This can be seen from \cref{lem:interval}.}
\end{lemma}
